Measurement Clusterization in Bayesian D-optimal Designs in Infinite Dimensions

### 2024-07-31 latex/flat.tex
Intro ready

\usepackage{thm-restate}
\usepackage{tikz,pgfplotstable}
\usetikzlibrary{patterns}
\pgfplotsset{compat=1.9} 
\newcommand{\R}{\ensuremath{\mathbb{R}}}
\newcommand{\eps}{\varepsilon}
\newcommand{\M}{\mathcal{M}}

@@

 \theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{conjecture}[theorem]{Conjecture}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
 \pubyear{2021}
\volume{TBA}
\issue{TBA}
\firstpage{1}
\lastpage{1}

@@

In this study, we focus on the Bayesian D-optimality
criterion. Bayesian D-optimality carries a simple and intuitive
meaning: measurements chosen according to the D-optimality criterion
maximize the expected Kullback-Liebler (KL) divergence from posterior
to prior \cite{chaloner1995, AlexanderianGloorGhattas14,
  CoverThomas91}. Utilizing KL divergence for quantifying differences
between distributions is standard practice and we will not motivate it
further here. For a discrete parameter $\param$ and data $\data$, the
KL divergence is defined as
\begin{equation}\label{eq:basic KL}
  D_{KL}\left (\Pr(\param|\data)||\Pr(\param)\right ) =  \sum_{\param} \log
  \frac{\Pr(\param|\data)}{\Pr(\param)} \Pr(\param|\data). 
\end{equation}
Of course, data $\data$ is not known before the experiment is
conducted, so we average over it to define the D-optimality criterion
as:
\begin{equation*}
  \mathbb{E}_{\data}\left [ D_{KL}\left (\Pr(\param|\data)||\Pr(\param)\right ) \right ]
\end{equation*}
We refer to a set of measuremnts that maximizes the D-optimality
criterion as a \emph{D-optimal design}.

For a linear model with Gaussian prior in finite dimensions, a
D-optimal design minimizes the determinant of the posterior covariance
\cite{chaloner1995}. In Section \ref{subsec:D optimal design} we give
a more general definition of the D-optimality criterion that applies
to arbitrary measures. We also show how the D-optimality criterion
generalizes to linear models over infinite-dimensional Hilbert spaces
(i.e.~function spaces).

@@

Solving the 1D heat equation is straightforward. Any initial heat
distribution $u_0\in L^2([0,1])$ that satisfies the homogeneous
Dirichlet boundary condition is a linear combination of sines
$\ev_n(x) = \sin(2\pi n x)$. Now, note that $\ev_n$ are in fact
eigenvectors of the Laplacian: $\Delta \ev_n = -(2\pi n)^2\ev_n$. The
linearity of the Laplacian implies the linearity of the heat equation
and we can solve in Fourier / frequency space. It is important to note
that when solving the heat equation we \emph{invert} the
Laplacian. I.e.~eigenvalues of the linear operator that describes the
time evolution of heat are $(2\pi n)^{-2}$.

@@

The inverse problem of the 1D heat equation is difficult ("ill-posed")
since heat spreads in a diffusive manner and any roughness in the
initial condition is quickly smoothed, see Supplementary movies S1 and
S2 (code generating these movies is located in module
\texttt{movies.py} in the accompanying
\href{https://github.com/yairdaon/OED}{repository}). For ill-posed
problems regularization is required, and we implement such
regularization via a fully Bayesian formulation of the inverse
problem.

Specifically, we set a prior on the initial condition $u_0 \sim
\mathcal{N}(0, (-\Delta)^{-1})$, where we impose the homogeneous
Dirichlet boundary condition on the Laplacian. We choose
$(-\Delta)^{-1}$ as our covariance operator following \cite{Stuart10}:
our choice ensures the power of the negative Laplacian inside the
parentheses is larger than $d/2$, where $d=1$ is the dimension of our
computational domain $\Omega$. One can intuitively understand this
choice from Weyl's asymptotics, that states that the eigenvalues of
Laplacian increase as $n^{d/2}$. Thus, our choice of covariance
$(-\Delta)^{-1}$ ensures higher modes decay sufficiently quickly, so
prior samples have sufficient regularity. Our choice of prior
covariance ensures the posterior is well-defined and well-behaved
\cite[Theorem 3.1 and Lemma 6.25]{Stuart10}.

\subsection{Measurement Clusterization}
Surprisingly, A- and D-optimal designs for inverse problems have been
observed to yield remarkably similar measurements in certain cases
\cite{fedorovDesignSpatialExperiments1996, nyberg2012, fedorov1997,
  Ucinski05, neitzel2019sparse}. This phenomenon is illustrated for
our toy inverse problem in Fig.~\ref{fig:clusterization illustration},
where D-optimal measurement locations are shown for different numbers
of measurements. Notably, for six measurements, a D-optimal design
yields two sets of measurements that are identical. Following
\cite{Ucinski05}, we refer to this intriguing phenomenon as
\emph{measurement clusterization}. We consider a design to be
\emph{clustered} when two or more of its constituent measurements are
identical.

\begin{figure}
    \centering
    \includegraphics[height=0.5\textwidth]{figs/dst_modelError0.png}
    \caption{Measurement clusterization for D-optimal designs for the
      inverse problem of the heat equation. Measurement locations were
      chosen according to the Bayesian D-optimality criterion of
      Theorem \ref{thm:d optimality}. Measurement locations are
      plotted over the computational domain \(\Omega = [0, 1]\)
      (x-axis), for varying numbers of measurements (y-axis). The
      colored numbers are measurement indices, plotted for visual
      clarity. Measurement clusterization already occurs for three
      measurements: the second measurement (red) is overlaid on the
      third (green). For five measurements, first (blue) and second
      (red) measurements are clustered, as well as the fourth (black)
      and the fifth (magenta).}
  \label{fig:clusterization illustration}
\end{figure}

@@

\subsection{Contribution}
The primary objective of this study is to provide a deep understanding
of measurement clusterization by addressing the aforementioned
questions. Our focus centers around investigating the Bayesian
D-optimality criterion. We conduct an analysis of Bayesian D-optimal
designs within the context of linear inverse problems over Hilbert
spaces and study two inverse problems: (a) In Sections
\ref{section:prelim} and \ref{section:D and grad} we propose a novel
relaxed model for an inverse problem where D-optimality maintains
analytical tractability, where D-optimal designs are identified via
Lagrange multipliers. This analytical framework facilitates the
exploration of the questions posed at the end of the previous
paragraph. We also study (b) the inverse problem of the 1D heat
equation from Section \ref{subsec:toy} above. Investigating both
inverse problems allows us to answer the questions posed in the
previous section:

\begin{enumerate}
\item \label{q:generic} \textbf{Is measurement clusterization a
  generic phenomenon?}
We give two complementing answers to this question. First, from a
  theoretical perspective, we show that clusterization mainly depends
  on how quickly the eigenvalues of the prior measure in observation
  space decay. This decay depends mostly on how ill-posed the problem
  at hand is and does not depend much on the prior. See Section
  \ref{section:vanishing} --- particularly, Theorem \ref{thm:char} and
  the discussion following it. We also show results of numerical
  experiments, where simulations of our model give rise to D-optimal
  designs that exhibit clusterization with high probability in Section
  \ref{subsec:lemma sims}. Thus, given the genericity of our model, we
  expect measurement clusterization to be a generic and ubiquitous
  phenomenon.

\item \label{q:mitigate} \textbf{Why does imposing correlations
  between observations alleviate measurement clusterization?} In
  Section \ref{section:non vanishing}, we rigorously demonstrate the
  role of model error in mitigating clusterization, thereby
  corroborating earlier observations made by various
  researchers. Specifically, our proof shows that identical
  measurements result in no gain in design criterion when observation
  error tends to zero. Moreover, in Section \ref{subsec:corr errors
    sims}, we show that an error term corresponding to correlations
  between measurements mitigates clusterization in the inverse problem
  of the 1D heat equation.

@@

\subsection{Forward Problems}\label{subsec:abstract OED}
In this section we give definitions and notations for forward problems,
which are an essential part of the inverse problems we discuss
later. Consider a "parameter space" \(\hilp\) and an "observation
space" \(\hilo\) --- both separable Hilbert spaces (the subscripts p
and o are for "parameter" and "observation", respectively). The
parameter space includes the quantity we seek to infer; in the inverse
problem of the heat equation, the parameter space $\hilp$ is where the
initial condition lives. The observation space $\hilo$, on the other
hand, is the space from which we take measurements; in the example of
the 1D heat equation, $u(\cdot, T) \in \hilo$.

The connection between parameter and observation spaces is made by the
\emph{forward operator} \(\fwd: \hilp \to \hilo\). We assume the
forward operator \(\fwd\) is linear. In the inverse problem of the 1D
heat, the forward operator is determined by the time evolution of the
1D heat equation \eqref{eq:heat1} and \eqref{eq:heat2}.

@@

Data is acquired via noisy observations, and we consider two types of
error terms: Spatially correlated model error \(\eps' \sim
\normal(0,\modcov)\) with \(\modcov\) a covariance operator; and
observation error denoted \(\eps \sim \normal(0, \sigma^2 I_m)\), with
\(I_m \in \mathbb{R}^{m \times m}\) the identity. Both error terms and
the prior (see below) are assumed independent of each other. Thus,
data is acquired via
\begin{align}\label{eq:inverse problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps.
\end{align}

@@

\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps) (\obs \eps' +
      \eps)^t ]
= \obs \modcov \obs^* + \sigma^2I_m , 
  \end{split}
\end{align}
where
\begin{align}\label{eq:modcov explained}
  \begin{split}
    [\obs \modcov \obs^*]_{ij} & = e_i^t \obs \modcov \obs^* e_j 
= \meas_i (\modcov \meas_j).\end{split}
\end{align}
Taking \(\modcov = 0\) is a common practice
\cite{tarantola2005,kaipio2005,Vogel02} and then \(\Sigma =
\sigma^2I_m\) is a scalar matrix which does not depend on \(\obs\).

\subsection{Bayesian Linear Inverse Problems}\label{subsec:bayesian inverse problems}
In the previous section, we saw how a parameter $u\in \hilp$ is
transported to the observation space via the forward operaror $\fwd u
\in \hilo$, how observations are generated from a parameter via $\obs
\fwd u$ and how observations and noise give rise to data $\data$. It
is time to formulate the process of inferring the parameter as a
Bayesian inverse problem. The first ingredient we need is a prior.

We take a Gaussian prior \(\param \sim \pr = \normal(\prmean
,\prcov)\) with some appropriate covariance operator \(\prcov\) on
\(\hilp\) \cite{Stuart10}. For example, for the inverse problem of the
1D heat equation we chose $\prcov = -\Delta^{-1}$ --- see Section
\ref{subsec:toy}. Note that \(\fwd \prcov \fwd^*\) is the prior
covariance in \(\hilo\) \cite{Stuart10}, and as such is assumed
invertible --- an assumption which we will use later (if \(\fwd\) has
a nontrivial kernel we utilize Occam's Razor and ignore said kernel
altogether).

@@

\subsection{Bayesian D-Optimal Designs}\label{subsec:D optimal design} 
A Bayesian D-optimal design maximizes the expected KL divergence
between posterior \(\post\) and prior measures \(\pr\). For arbitrary
posterior and prior measures, the KL divergence is defined,
analogously to eq.~\ref{eq:basic KL}, via the Radon-Nikodym
derivative:
\begin{equation*}
  D_{KL}(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
\end{equation*}

@@

\begin{theorem}[Alexanderian, Gloor, Ghattas \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let \(\pr = \normal(\prmean,\prcov)\) be a Gaussian prior on \(\hilp\)
  and let \(\post = \normal(\postmean,\postcov)\) the posterior measure
  on \(\hilp\) for the Bayesian linear inverse problem \(\data = \obs
  \fwd\param + \obs \eps' + \eps\) discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
&= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}

@@

\begin{definition}\label{def:d optimality}
  We say \(\opt\) is \emph{D-optimal} if \(\opt =
  \argmax_{\obs} \tar(\obs)\), where entries of \(\obs \in (\hilo^*)^m\)
  are constrained to some allowed set of measurements in \(\hilo^*\).
\end{definition}

Intuition for Theorem \ref{thm:d optimality} can be gained by
recalling from Section \ref{subsec:D} that for a Bayesian linear model
in finite dimensions, with Gaussian prior and Gaussian noise, a
D-optimal design minimizes the determinant of the posterior covariance
matrix. Theorem
\ref{thm:d optimality} and Definition \ref{def:d optimality} carry a
similar intuition:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \frac12 \log \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \text{ by \eqref{eq:objective}}\\
    &= \frac12 \log \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \frac12 \log \det \prcov \postcov^{-1} \text{ by \eqref{eq:postcov}}.
\end{split}
\end{align*}
We think of \(\prcov\) as constant, in the sense that $\prcov$ does
not depend on data $\data$. Thus, a D-optimal design minimizes a
quantity analogous to the posterior covariance determinant, similarly
to the finite-dimensional case.

 \section{The Constrained Optimization Problem of D-Optimal Design}\label{section:D and grad}
We seek a formulation of the D-optimal design problem via Lagrange
multipliers. We first find the gradient of $\tar$, then we suggest
unit-norm constraints on $\obs$ and find their gradients. Results of
this section are summarized in Theorem \ref{thm:constrained}. First,
recall that:
\begin{definition}\label{def:var}
  Let $F$ a real valued function of $\obs$. The first variation of $F$
  at $\obs$ in the direction $V$ is:
  \begin{equation*}
    \delta F(\obs) V := \frac{\der}{\der \tau}\Big |_{\tau=0}  F( \obs + \tau V).
  \end{equation*}

@@

\begin{proposition}\label{prop:tar grad}
  The gradient of the D-optimality objective $\tar$ is
  \begin{equation*}
\nabla \tar(\obs) = ( I - \modcov \obs^* \Sigma^{-1}\obs ) \fwd
    \postcov \fwd^* \obs^* \Sigma^{-1}
  \end{equation*}
\end{proposition}

@@

\begin{proposition}\label{prop:bigger better}
  Let $\obs = (\meas_1,\dots,\meas_m)^t$, $j \in \{1,\dots,m\}$,
  $\sigma^2 > 0$ and $|\zeta| > 1$. Then $\tar(\obs)$ increases if we
  use $\zeta \meas_j$ in $\obs$ instead of $\meas_j$.
\end{proposition}

\begin{proof} 
  Fix $j=1,\dots,m$ and take $V:= e_j e_j^t \obs$. For $u
  \in \hilo$:
  \begin{equation*}
    Vu = e_je_j^t (\meas_1(u),\dots,\meas_m(u) )^t = e_j \meas_j(u)
    = (0,\dots,0,\meas_j(u),0,\dots,0)^t.
  \end{equation*}
We now calculate the variation of $\tar$ at $\obs$ in the direction
  of $V$. Denote $\tmp: = \fwd \postcov \fwd^*$. From Proposition
  \ref{prop:tar grad}:
  \begin{align*}
     \delta \tar(\obs) V 
    &= \tr{V ( I - \modcov \obs^*\Sigma^{-1}\obs) \tmp \obs^* \Sigma^{-1}} \\
&= \tr{e_je_j^t \obs ( I - \modcov \obs^*\Sigma^{-1}\obs) \tmp \obs^* \Sigma^{-1}} \\
&= e_j^t \obs ( I - \modcov \obs^*\Sigma^{-1}\obs) \tmp \obs^* \Sigma^{-1}e_j \\
&= e_j^t ( I - \obs \modcov \obs^*\Sigma^{-1})\obs \tmp \obs^* \Sigma^{-1}e_j \\  
&=  e_j^t(\Sigma-\obs \modcov \obs^*) \Sigma^{-1}\obs \tmp \obs^* \Sigma^{-1}e_j \\
&=\sigma^2 e_j^t \Sigma^{-1}\obs \tmp \obs^* \Sigma^{-1}e_j
    \text{ by \eqref{eq:Sigma} }\\
&=\sigma^2 e_j^t \Sigma^{-1}\obs \fwd \postcov \fwd^* \obs^* \Sigma^{-1}e_j.
  \end{align*} 
  Since $\postcov$ is positive definite, we conclude that $\delta
  \tar(\obs) V > 0$. This means that increasing the magnitude of the
  $j^{\text{th}}$ measurement functional increases $\tar(\obs)$.
\end{proof}

Proposition \ref{prop:bigger better} implies that it is a good idea to
bound the norm of measurements. If, for example, we can take
measurements in $\textup{span}\{\meas\}$ for some $\meas \neq 0$, then
the the D-optimality criterion is unbounded. In contrast, in any
real-life problem where sensors are concerned, the norm of
measurements recorded by sensors is always one\footnote{Again, our
analysis does not apply to point evaluations. These are just used for
motivation and can be approximated by elements in $\hilo^*$ as long as
$\hilo$ is a function space, e.g.~$L^2(\Omega)$.}:

>>>>

\usepackage{thm-restate}
\usepackage{tikz,pgfplotstable}
\usetikzlibrary{patterns}
\pgfplotsset{compat=1.9} \usepackage{xcolor}

\newcommand{\R}{\ensuremath{\mathbb{R}}}
\newcommand{\eps}{\varepsilon}
\newcommand{\M}{\mathcal{M}}

@@

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{conjecture}[theorem]{Conjecture}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\pubyear{2021}
\volume{TBA}
\issue{TBA}
\firstpage{1}
\lastpage{1}

@@

In this study, we focus on the Bayesian D-optimality
criterion. Bayesian D-optimality carries a simple and intuitive
meaning: measurements chosen according to the D-optimality criterion
maximize the expected Kullback-Liebler (KL) divergence from posterior
to prior \cite{chaloner1995, AlexanderianGloorGhattas14,
  CoverThomas91}. Utilizing KL divergence for quantifying differences
between distributions is standard practice and we will not motivate it
further here. For a discrete parameter $\param$ and data $\data$, the
KL divergence is defined as
\begin{equation}\label{eq:basic_KL}
  D_{KL}\left (\Pr(\param|\data)||\Pr(\param)\right ) =  \sum_{\param} \log
  \frac{\Pr(\param|\data)}{\Pr(\param)} \Pr(\param|\data). 
\end{equation}
Of course, data $\data$ is not known before the experiment is
conducted, so we average over it to define the D-optimality criterion
as:
\begin{equation*}
  \mathbb{E}_{\data}\left [ D_{KL}\left (\Pr(\param|\data)||\Pr(\param)\right ) \right ]
\end{equation*}
We refer to a set of measuremnts that maximizes the D-optimality
criterion as a \emph{D-optimal design}.

For a linear model with Gaussian prior in finite dimensions, a
D-optimal design minimizes the determinant of the posterior covariance
\cite{chaloner1995}. In Section \ref{subsec:D_optimal_design} we give
a more general definition of the D-optimality criterion that applies
to arbitrary measures. We also show how the D-optimality criterion
generalizes to linear models over infinite-dimensional Hilbert spaces
(i.e.~function spaces).

@@

Solving the 1D heat equation is straightforward. Any initial heat
distribution $u_0\in L^2([0,1])$ that satisfies the homogeneous
Dirichlet boundary condition eq.~\eqref{eq:heat2} is a linear
combination of sines $\ev_n(x) = \sin(\pi n x)$, so $u_0 = \sum_{n\geq
  1} a_n \ev_n$ for some $\{a_n\}_{n\geq 1}$. Now, note that $\ev_n$
are in fact eigenvectors of the Laplacian: $\Delta \ev_n = -\pi^2
n^2\ev_n$. It should not be hard to believe that eq.~\eqref{eq:heat1}
implies $u(\cdot, T) = \sum_{n\geq 1} a_n \exp(-\pi^2 n^2T ) \ev_n$
\cite{renardy2006PDE}. Thus, eigenvectors of the linear operator that
describes the time evolution of heat are $\ev_n(x) = \sin(\pi n x)$
with corresponding eigenvalues
\begin{equation}\label{eq:decay}
  \exp(-\pi^2 n^2T ).
\end{equation}

@@

The inverse problem of the 1D heat equation is difficult ("ill-posed")
since heat spreads in a diffusive manner and any roughness in the
initial condition is quickly smoothed, as implied by the squared
exponential decay as $n\to\infty$ in eq.~\eqref{eq:decay}. See
Supplementary movies S1 and S2 for an illustration of this phenomenon
(code generating these movies is located in module \texttt{movies.py}
in the accompanying
\href{https://github.com/yairdaon/OED}{repository}). For ill-posed
problems regularization is required, and we implement such
regularization via a fully Bayesian formulation of the inverse
problem.

We choose a prior for the initial condition $u_0 \sim \mathcal{N}(0,
(-\Delta)^{-1})$, with homogeneous Dirichlet boundary condition. Choosing a prior This
choice ensures the posterior is well-defined and prior realizations
are well-behaved, see \cite[Theorem 3.1 and Lemma 6.25]{Stuart10} for
details.

\subsection{Measurement Clusterization}
Surprisingly, A- and D-optimal designs for inverse problems have been
observed to yield remarkably similar measurements in certain cases
\cite{fedorovDesignSpatialExperiments1996, nyberg2012, fedorov1997,
  Ucinski05, neitzel2019sparse}. This phenomenon is illustrated for
our toy inverse problem in Fig.~\ref{fig:clusterization_illustration},
where D-optimal measurement locations are shown for different numbers
of measurements. Notably, for six measurements, a D-optimal design
yields two sets of measurements that are identical. Following
\cite{Ucinski05}, we refer to this intriguing phenomenon as
\emph{measurement clusterization}. We consider a design to be
\emph{clustered} when two or more of its constituent measurements are
identical.

\begin{figure}
    \centering
    \includegraphics[height=0.5\textwidth]{figs/dst_modelError0.png}
    \caption{Measurement clusterization for D-optimal designs for the
      inverse problem of the heat equation. Measurement locations were
      chosen according to the Bayesian D-optimality criterion of
      Theorem \ref{thm:d_optimality}. Measurement locations are
      plotted over the computational domain \(\Omega = [0, 1]\)
      (x-axis), for varying numbers of measurements (y-axis). The
      colored numbers are measurement indices, plotted for visual
      clarity. Measurement clusterization already occurs for three
      measurements: the second measurement (red) is overlaid on the
      third (green). For five measurements, first (blue) and second
      (red) measurements are clustered, as well as the fourth (black)
      and the fifth (magenta).}
  \label{fig:clusterization_illustration}
\end{figure}

@@

\subsection{Contribution}
The primary objective of this study is to provide a deep understanding
of measurement clusterization by addressing the aforementioned
questions. Our focus centers around investigating the Bayesian
D-optimality criterion. We conduct an analysis of Bayesian D-optimal
designs within the context of linear inverse problems over Hilbert
spaces and study two inverse problems: (a) In Sections
\ref{section:prelim} and \ref{section:D_and_grad} we propose a novel
relaxed model for an inverse problem where D-optimality maintains
analytical tractability, where D-optimal designs are identified via
Lagrange multipliers. This analytical framework facilitates the
exploration of the questions posed at the end of the previous
paragraph. We also study (b) the inverse problem of the 1D heat
equation from Section \ref{subsec:toy} above. Investigating both
inverse problems allows us to answer the questions posed in the
previous section:

\begin{enumerate}
\item \label{q:generic} \textbf{Is measurement clusterization a
  generic phenomenon?}
We give two complementing answers to this question. First, from a
  theoretical perspective, we show that clusterization mainly depends
  on how quickly the eigenvalues of the prior measure in observation
  space decay. This decay depends mostly on how ill-posed the problem
  at hand is and does not depend much on the prior. See Section
  \ref{section:vanishing} --- particularly, Theorem \ref{thm:char} and
  the discussion following it. We also show results of numerical
  experiments, where simulations of our model give rise to D-optimal
  designs that exhibit clusterization with high probability in Section
  \ref{subsec:lemma_sims}. Thus, given the genericity of our model, we
  expect measurement clusterization to be a generic and ubiquitous
  phenomenon.

\item \label{q:mitigate} \textbf{Why does imposing correlations
  between observations alleviate measurement clusterization?} In
  Section \ref{section:non_vanishing}, we rigorously demonstrate the
  role of model error in mitigating clusterization, thereby
  corroborating earlier observations made by various
  researchers. Specifically, our proof shows that identical
  measurements result in no gain in design criterion when observation
  error tends to zero. Moreover, in Section
  \ref{subsec:corr_errors_sims}, we show that an error term
  corresponding to correlations between measurements mitigates
  clusterization in the inverse problem of the 1D heat equation.

@@

\subsection{Forward Problems}\label{subsec:abstract_OED}
In this section we give definitions and notations for forward problems,
which are an essential part of the inverse problems we discuss
later. Consider a "parameter space" \(\hilp\) and an "observation
space" \(\hilo\) --- both separable Hilbert spaces (the subscripts p
and o are for "parameter" and "observation", respectively). The
parameter space includes the quantity we seek to infer; in the inverse
problem of the heat equation, the parameter space $\hilp$ is where the
initial condition lives. The observation space $\hilo$, on the other
hand, is the space from which we take measurements; in the example of
the 1D heat equation, $u(\cdot, T) \in \hilo$.

The connection between parameter and observation spaces is made by the
\emph{forward operator} \(\fwd: \hilp \to \hilo\). We assume the
forward operator \(\fwd\) is linear. In the inverse problem of the 1D
heat equatin, the forward operator is determined by the time evolution
of the 1D heat equation \eqref{eq:heat1} and \eqref{eq:heat2}.

@@

Data is acquired via noisy observations, and we consider two types of
error terms: Spatially correlated model error \(\eps' \sim
\normal(0,\modcov)\) with \(\modcov\) a covariance operator; and
observation error denoted \(\eps \sim \normal(0, \sigma^2 I_m)\), with
\(I_m \in \mathbb{R}^{m \times m}\) the identity. Both error terms and
the prior (see below) are assumed independent of each other. Thus,
data is acquired via
\begin{align}\label{eq:inverse_problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps.
\end{align}

@@

\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps) (\obs \eps' +
      \eps)^t ]
= \obs \modcov \obs^* + \sigma^2I_m , 
  \end{split}
\end{align}
where
\begin{align}\label{eq:modcov_explained}
  \begin{split}
    [\obs \modcov \obs^*]_{ij} & = e_i^t \obs \modcov \obs^* e_j 
= \meas_i (\modcov \meas_j).\end{split}
\end{align}
Taking \(\modcov = 0\) is a common practice
\cite{tarantola2005,kaipio2005,Vogel02} and then \(\Sigma =
\sigma^2I_m\) is a scalar matrix which does not depend on \(\obs\).

\subsection{Bayesian Linear Inverse Problems}\label{subsec:bayesian_inverse_problems}
In the previous section, we saw how a parameter $u\in \hilp$ is
transported to the observation space via the forward operaror $\fwd u
\in \hilo$, how observations are generated from a parameter via $\obs
\fwd u$ and how observations and noise give rise to data $\data$. It
is time to formulate the process of inferring the parameter as a
Bayesian inverse problem. The first ingredient we need is a prior.

We take a Gaussian prior \(\param \sim \pr = \normal(\prmean
,\prcov)\) with some appropriate covariance operator \(\prcov\) on
\(\hilp\) \cite{Stuart10}. For example, for the inverse problem of the
1D heat equation we chose $\prcov = (-\Delta)^{-1}$ --- see Section
\ref{subsec:toy}. Note that \(\fwd \prcov \fwd^*\) is the prior
covariance in \(\hilo\) \cite{Stuart10}, and as such is assumed
invertible --- an assumption which we will use later (if \(\fwd\) has
a nontrivial kernel we utilize Occam's Razor and ignore said kernel
altogether).

@@

\subsection{Bayesian D-Optimal Designs}\label{subsec:D_optimal_design} 
A Bayesian D-optimal design maximizes the expected KL divergence
between posterior \(\post\) and prior measures \(\pr\). For arbitrary
posterior and prior measures, the KL divergence is defined,
analogously to eq.~\ref{eq:basic_KL}, via the Radon-Nikodym
derivative:
\begin{equation*}
  D_{KL}(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
\end{equation*}

@@

\begin{theorem}[Alexanderian, Gloor, Ghattas \cite{AlexanderianGloorGhattas14}]\label{thm:d_optimality}
  Let \(\pr = \normal(\prmean,\prcov)\) be a Gaussian prior on \(\hilp\)
  and let \(\post = \normal(\postmean,\postcov)\) the posterior measure
  on \(\hilp\) for the Bayesian linear inverse problem \(\data = \obs
  \fwd\param + \obs \eps' + \eps\) discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
&= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}

@@

\begin{definition}\label{def:d_optimality}
  We say \(\opt\) is \emph{D-optimal} if \(\opt =
  \argmax_{\obs} \tar(\obs)\), where entries of \(\obs \in (\hilo^*)^m\)
  are constrained to some allowed set of measurements in \(\hilo^*\).
\end{definition}

Intuition for Theorem \ref{thm:d_optimality} can be gained by
recalling from Section \ref{subsec:D} that for a Bayesian linear model
in finite dimensions, with Gaussian prior and Gaussian noise, a
D-optimal design minimizes the determinant of the posterior covariance
matrix. Theorem
\ref{thm:d_optimality} and Definition \ref{def:d_optimality} carry a
similar intuition:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \frac12 \log \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \text{ by \eqref{eq:objective}}\\
    &= \frac12 \log \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \frac12 \log \det \prcov \postcov^{-1} \text{ by \eqref{eq:postcov}}.
\end{split}
\end{align*}
We think of \(\prcov\) as constant, in the sense that $\prcov$ does
not depend on data $\data$. Thus, a D-optimal design minimizes a
quantity analogous to the posterior covariance determinant, similarly
to the finite-dimensional case.

 \section{The Constrained Optimization Problem of D-Optimal Design}\label{section:D_and_grad}
We seek a formulation of the D-optimal design problem via Lagrange
multipliers. We first find the gradient of $\tar$, then we suggest
unit-norm constraints on $\obs$ and find their gradients. Results of
this section are summarized in Theorem \ref{thm:constrained}. First,
recall that:
\begin{definition}\label{def:var}
  Let $F$ a real valued function of $\obs$. The first variation of $F$
  at $\obs$ in the direction $V$ is:
  \begin{equation*}
    \delta F(\obs) V := \frac{\der}{\der \tau}\Big |_{\tau=0}  F( \obs + \tau V).
  \end{equation*}

@@

\begin{proposition}\label{prop:tar_grad}
  The gradient of the D-optimality objective $\tar$ is
  \begin{equation*}
\nabla \tar(\obs) = ( I - \modcov \obs^* \Sigma^{-1}\obs ) \fwd
    \postcov \fwd^* \obs^* \Sigma^{-1}
  \end{equation*}
\end{proposition}

@@

\begin{proposition}\label{prop:bigger_better}
  Let $\obs = (\meas_1,\dots,\meas_m)^t$, $j \in \{1,\dots,m\}$,
  $\sigma^2 > 0$ and $|\zeta| > 1$. Then $\tar(\obs)$ increases if we
  use $\zeta \meas_j$ in $\obs$ instead of $\meas_j$.
\end{proposition}

\begin{proof} 
  Fix $j=1,\dots,m$ and take $V:= e_j e_j^t \obs$. For $u
  \in \hilo$:
  \begin{equation*}
    Vu = e_je_j^t (\meas_1(u),\dots,\meas_m(u) )^t = e_j \meas_j(u)
    = (0,\dots,0,\meas_j(u),0,\dots,0)^t.
  \end{equation*}
We now calculate the variation of $\tar$ at $\obs$ in the direction
  of $V$. Denote $\tmp: = \fwd \postcov \fwd^*$. From Proposition
  \ref{prop:tar_grad}:
  \begin{align*}
     \delta \tar(\obs) V 
    &= \tr{V ( I - \modcov \obs^*\Sigma^{-1}\obs) \tmp \obs^* \Sigma^{-1}} \\
&= \tr{e_je_j^t \obs ( I - \modcov \obs^*\Sigma^{-1}\obs) \tmp \obs^* \Sigma^{-1}} \\
&= e_j^t \obs ( I - \modcov \obs^*\Sigma^{-1}\obs) \tmp \obs^* \Sigma^{-1}e_j \\
&= e_j^t ( I - \obs \modcov \obs^*\Sigma^{-1})\obs \tmp \obs^* \Sigma^{-1}e_j \\  
&=  e_j^t(\Sigma-\obs \modcov \obs^*) \Sigma^{-1}\obs \tmp \obs^* \Sigma^{-1}e_j \\
&=\sigma^2 e_j^t \Sigma^{-1}\obs \tmp \obs^* \Sigma^{-1}e_j
    \text{ by \eqref{eq:Sigma} }\\
&=\sigma^2 e_j^t \Sigma^{-1}\obs \fwd \postcov \fwd^* \obs^* \Sigma^{-1}e_j.
  \end{align*} 
  Since $\postcov$ is positive definite, we conclude that $\delta
  \tar(\obs) V > 0$. This means that increasing the magnitude of the
  $j^{\text{th}}$ measurement functional increases $\tar(\obs)$.
\end{proof}

Proposition \ref{prop:bigger_better} implies that it is a good idea to
bound the norm of measurements. If, for example, we can take
measurements in $\textup{span}\{\meas\}$ for some $\meas \neq 0$, then
the D-optimality criterion is unbounded. In contrast, in any real-life
problem where sensors are concerned, the norm of measurements recorded
by sensors is always one\footnote{Again, our analysis does not apply
to point evaluations. These are just used for motivation and can be
approximated by elements in $\hilo^*$ as long as $\hilo$ is a function
space, e.g.~$L^2(\Omega)$.}:
